manuscript_figures

### 2024-07-05 scripts/manuscript_figures.Rmd
updated figures and models

# rerun model with log and without outliers
lmm_model <- lmer(log(mean_fish_biomass + 1.1) ~ log(mean_kelp_biomass + 1.1) * YEAR + (1 | SITE), data = data_clean)
#check normality
qqPlot(residuals(lmm_model)) #improved to normal

@@

# Export to CSV file
write.csv(lmm_tidy, "output/lmm_model_results.csv", row.names = FALSE)

@@

# Create scatter plot
figure.3 <- ggplot(aggregated.data.morph, aes(x = log(mean_kelp_biomass + 1.1), y = log(mean_fish_biomass + 1.1))) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, formula = y ~ x, color = "blue") +

@@

  # p-value
  annotate("text", x = Inf, y = -Inf, 
           label = paste0("p = ", format(p_value_kelp_biomass, scientific = FALSE, digits = 2)), 
           hjust = 1.3, vjust = -38.5, size = 4) +

  labs(x = expression("Log Kelp Biomass (g/" ~ m^2 ~ ")"), y = expression("Log Fish Biomass (g/" ~ m^2 ~ ")")) +
  theme_classic() +
  theme(plot.margin = margin(t = 10, r = 30, b = 10, l = 10, unit = "pt")) +
  coord_cartesian(clip = 'off') +
  theme(
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18));figure.3

# Save the plot to a file
ggsave("output/figures/figure_3.png", plot = figure.3, width = 8, height = 6)
```

@@

# Time series plot for kelp biomass with standard error bars from 2016-2023
kelp.timeseries.2016 <- ggplot(aggregated.kelp.morph, aes(x = YEAR, y = log_mean_kelp_biomass, shape = SITE)) +
  geom_line(aes(group = SITE), color = "black") +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = log_mean_kelp_biomass - log_se_kelp_biomass, 
                    ymax = log_mean_kelp_biomass + log_se_kelp_biomass), 
                width = 0.2) +
  scale_shape_manual(values = shape_values) +
  labs(x = "Year", y = expression("Log Mean Kelp Biomass (g/" ~ m^2 ~ ")"), shape = "Site") +
  theme_bw() +
  scale_x_continuous(breaks = unique(aggregated.kelp.morph$YEAR)) +
  theme(
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18)
  );kelp.timeseries.2016

@@

# Time series plot for kelp biomass with standard error bars for 2019-2023
kelp.timeseries.2019_2023 <- ggplot(filtered_data, aes(x = YEAR, y = log_mean_kelp_biomass, shape = SITE)) +
  geom_line(aes(group = SITE), color = "black") +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = log_mean_kelp_biomass - log_se_kelp_biomass, 
                    ymax = log_mean_kelp_biomass + log_se_kelp_biomass), 
                width = 0.2) +
  scale_shape_manual(values = shape_values) +
  labs(x = "Year", y = expression("Log Mean Kelp Biomass (g/" ~ m^2 ~ ")"), shape = "Site") +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18)
  );kelp.timeseries.2019_2023

@@

# Fit linear regression models for each SITE separately
slope_results <- morph.filt %>%
  mutate(log_avg_kelp_biomass = log(average_kelp_biomass + 1.1)) %>%
  group_by(SITE) %>%
  summarize(
    mean_slope = coef(lm(log_avg_kelp_biomass ~ YEAR))[2],  # Mean slope
    se_slope = summary(lm(log_avg_kelp_biomass ~ YEAR))$coefficients[2, "Std. Error"],  # Standard error of slope
    mean_log_avg_kelp_biomass = mean(log_avg_kelp_biomass)  # Mean log average kelp biomass
  )

@@

# Plotting the mean rates of change with error bars (standard error horizontally)
rate_of_change <- ggplot(slope_results, aes(x = rate_of_change_per_year, y = mean_log_avg_kelp_biomass, shape = SITE)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = rate_of_change_per_year - se_rate_of_change, xmax = rate_of_change_per_year + se_rate_of_change), height = 0.2) +
  labs(
    x = expression("Rate of Change ("~yr^-1 ~")"),
    y = expression("Log Mean Kelp Biomass (g/"~m^2~")"), shape = "Site") +
  theme_bw() +
  scale_shape_manual(values = shape_values, guide = guide_legend(title = "Site")) +   # Use shape values and customize legend
  theme(
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18));rate_of_change

@@

# Calculate fish biomass 'W' using the length-weight relationship
PB$biomass <- with(PB, coef.a * (actual_size)^coef.b)

@@

# Plotting nMDS results
nmds_plot <- ggplot(scores, aes(x = NMDS1, y = NMDS2, color = SITE, shape = as.factor(YEAR))) +
  geom_point(size = 8) +
  labs(x = "NMDS1", y = "NMDS2", color = "Site", shape = "Year") +
  stat_ellipse(aes(group = YEAR), level = 0.95, linetype = 2, size = 1, alpha = 0.2)+
  scale_color_manual(values = c("Fort Wetherill" = "grey", "King's Beach" = "black")) +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18));nmds_plot

@@

PERMANOVA and SIMPER analysis
```{r}
# Remove rows where all abundance values are zero
non_zero_rows <- rowSums(community_data) > 0
community_data <- community_data[non_zero_rows, ]
environmental_data <- environmental_data[non_zero_rows, ]

@@

# Create dataframe from PERMANOVA results
permanova_df <- data.frame(
  Factor = c("SITE", "YEAR", "Residual", "Total"),
  Df = c(1, 4, 37, 42),
  SumOfSqs = c(0.2632, 2.8619, 7.4000, 10.5251),
  R2 = c(0.02501, 0.27191, 0.70308, 1.00000),
  F = c(1.3160, 3.5774, NA, NA),
  Pr_F = c(0.235, 0.001, NA, NA)  # Changed column name to Pr_F
)

@@

# Customize the flextable
flex_perm <- flex_perm %>%
  set_header_labels(
    Factor = "Factor",
    Df = "Df",
    SumOfSqs = "Sum of Squares",
    R2 = "R2",
    F = "F",
    Pr_F = "Pr(>F)"  # Changed to match the new column name
  ) %>%
  colformat_double(digits = 4) %>%
  bold(i = ~ Factor == "YEAR", j = "Pr_F") %>%  # Changed to match the new column name
  add_header_lines(values = c(
    "Permutation test for adonis under reduced model",
    "Terms added sequentially (first to last)",
    "Permutation: free",
    "Number of permutations: 999"
  )) %>%
  add_footer_lines(c(
    "Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1",
    "adonis2(formula = community_data ~ SITE + YEAR, data = environmental_data, permutations = 999, method = \"bray\")"
  )) %>%
  autofit() %>%
  theme_box()

@@

# For the SIMPER analysis, we'll use the original (non-transformed) data, but still excluding the all-zero rows
simper_result <- simper(community_data, group = environmental_data$SITE, permutations = 999)

@@

@@

# Extract the relevant information from the SIMPER results
simper_data <- data.frame(
  Species = simper_result$`Fort Wetherill_King's Beach`$species,
  Contribution = simper_result$`Fort Wetherill_King's Beach`$average,
  Fort_Wetherill = simper_result$`Fort Wetherill_King's Beach`$ava,
  Kings_Beach = simper_result$`Fort Wetherill_King's Beach`$avb
)

@@

# Create the plot
ggplot(simper_long, aes(x = Species, y = Abundance, fill = Site)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(data = simper_data, 
            aes(x = Species, y = max(simper_long$Abundance) + 0.5, 
                label = sprintf("%.1f%%", Percentage)),
            position = position_dodge(width = 0.9),
            vjust = -0.5) +
  labs(title = "Species Contributions to Differences Between Sites",
       subtitle = "Based on SIMPER analysis",
       y = "Average Abundance",
       x = "Species") +
  scale_fill_manual(values = c("Fort_Wetherill" = "grey", "Kings_Beach" = "black")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.title = element_blank())

@@

PERMANOVA results:
adonis2(formula = community_data ~ SITE + YEAR, data = environmental_data, permutations = 999, method = "bray")
         Df SumOfSqs      R2      F Pr(>F)    
SITE      1   0.2632 0.02501 1.3160  0.235    
YEAR      4   2.8619 0.27191 3.5774  0.001 ***
Residual 37   7.4000 0.70308                  
Total    42  10.5251 1.00000

SIMPER results:
Contrast: Fort Wetherill_King's Beach 
            average      sd   ratio     ava     avb cumsum     p  
molluscs    0.56710 0.30146 1.88120 7.64700 5.84300  0.894 0.803  
echinoderms 0.04820 0.10148 0.47480 0.14700 0.34300  0.970 0.067 .
crustacean  0.01920 0.04652 0.41350 0.30000 0.03700  1.000 0.713  
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
Permutation: free
Number of permutations: 999

>>>>

# rerun model with log and without outliers
lmm_model <- lmer(log(mean_fish_biomass + 1.1) ~ log(mean_kelp_biomass + 1.1) + (1 | YEAR) + (1 | SITE), data = data_clean)
#check normality
qqPlot(residuals(lmm_model)) #improved to normal
summary(lmm_model)

# Save flextable as PNG
png_file <- "output/model.result.tables/rugosity.not.included/summary_fish_kelp_biomass.png"
save_as_image(flex, png_file)

@@

@@

# Create scatter plot
figure.3 <- ggplot(aggregated.data.morph, aes(mean_kelp_biomass, mean_fish_biomass)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, formula = y ~ x, color = "blue") +

@@

  # p-value
  annotate("text", x = Inf, y = -Inf, 
           label = paste0("p = ", format(p_value_kelp_biomass, scientific = FALSE, digits = 2)), 
           hjust = 1.3, vjust = -39, size = 4) +

  labs(x = expression("Kelp Biomass (g/" ~ m^2 ~ ")"), y = expression("Fish Biomass (g/" ~ m^2 ~ ")")) +
  theme_classic() +
  theme(plot.margin = margin(t = 10, r = 30, b = 10, l = 10, unit = "pt")) +
  coord_cartesian(clip = 'off') +
  theme(
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18));figure.3

# Save the plot to a file
ggsave("output/figures/figure_3.png", plot = figure.3, width = 8, height = 6)

@@

# Time series plot for kelp biomass with standard error bars from 2016-2023
kelp.timeseries.2016 <- ggplot(aggregated.kelp.morph, aes(x = YEAR, y = mean_kelp_biomass, shape = SITE)) +
  geom_line(aes(group = SITE), color = "black") +
  geom_point(size = 6) +
  geom_errorbar(aes(ymin = mean_kelp_biomass - se_kelp_biomass, 
                    ymax = mean_kelp_biomass + se_kelp_biomass), 
                width = 0.2) +
  scale_shape_manual(values = shape_values) +
  labs(x = "Year", y = expression("Mean Kelp Biomass (g/" ~ m^2 ~ ")"), shape = "Site") +
  theme_bw() +
  scale_x_continuous(breaks = unique(aggregated.kelp.morph$YEAR)) +
  theme(
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18),
    legend.position = c(1, 1),  # Positioning legend inside the plot area at the top right
    legend.justification = c(1, 1),  # Anchoring the legend to the top right corner
    legend.background = element_rect(fill = "white", color = "black")  # Optional: adding a background for better visibility
  );kelp.timeseries.2016

@@

# Time series plot for kelp biomass with standard error bars for 2019-2023
kelp.timeseries.2019_2023 <- ggplot(filtered_data, aes(x = YEAR, y = log_mean_kelp_biomass, shape = SITE)) +
  geom_line(aes(group = SITE), color = "black") +
  geom_point(size = 8) +
  geom_errorbar(aes(ymin = log_mean_kelp_biomass - log_se_kelp_biomass, 
                    ymax = log_mean_kelp_biomass + log_se_kelp_biomass), 
                width = 0.2) +
  scale_shape_manual(values = shape_values) +
  labs(x = "Year", y = expression("Log Mean Kelp Biomass (g/" ~ m^2 ~ ")"), shape = "Site") +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18)
  );kelp.timeseries.2019_2023

@@

# Fit linear regression models for each SITE separately
slope_results <- morph.filt %>%
  mutate(log_avg_kelp_biomass = log(average_kelp_biomass + 1.1)) %>%
  group_by(SITE) %>%
  summarize(
    mean_slope = coef(lm(log_avg_kelp_biomass ~ YEAR))[2],  # Mean slope
    se_slope = summary(lm(log_avg_kelp_biomass ~ YEAR))$coefficients[2, "Std. Error"],  # Standard error of slope
    mean_log_avg_kelp_biomass = mean(log_avg_kelp_biomass),  # Mean log average kelp biomass
    mean_nonlog_avg_kelp_biomass = mean(average_kelp_biomass)  # Mean non-log average kelp biomass
  )

@@

# Plotting the mean rates of change with error bars (standard error horizontally)
rate_of_change <- ggplot(slope_results, aes(x = rate_of_change_per_year, y = mean_nonlog_avg_kelp_biomass, shape = SITE)) +
  geom_point(size = 8) +
  geom_errorbarh(aes(xmin = rate_of_change_per_year - se_rate_of_change, xmax = rate_of_change_per_year + se_rate_of_change), height = 0.9) +
  labs(
    x = expression("Rate of Change ("~yr^-1 ~")"),
    y = expression("Mean Kelp Biomass (g/"~m^2~")"), shape = "Site") +
  theme_bw() +
  scale_shape_manual(values = shape_values, guide = guide_legend(title = "Site")) +   # Use shape values and customize legend
  theme(
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18), ,
    legend.position = "none");rate_of_change

@@

#filter out data from 2016 and 2017 due to size bins
PB_filtered <- PB %>%
  filter(YEAR != 2016 & YEAR != 2017)

# Calculate fish biomass 'W' using the length-weight relationship
PB_filtered$biomass <- with(PB, coef.a * (actual_size)^coef.b)

@@

# Define custom shape values for more than 8 categories
custom_shape_values <- c(16, 17, 18, 19, 20, 21, 22, 23)

# Use these custom shapes in your plot
nmds_plot <- ggplot(scores, aes(x = NMDS1, y = NMDS2, color = SITE, shape = as.factor(YEAR))) +
  geom_point(size = 8) +
  #stat_ellipse(aes(group = YEAR), level = 0.95, linetype = 2, size = 1, alpha = 0.5) + # Add ellipses
  labs(x = "NMDS1", y = "NMDS2", color = "Site", shape = "Year") +
  scale_color_manual(values = c("Fort Wetherill" = "grey", "King's Beach" = "black")) +
  scale_shape_manual(values = custom_shape_values) +  # Specify the custom shapes
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18)
  );nmds_plot

@@

PERMANOVA and SIMPER analysis
```{r}
# Ensure only numeric columns are selected for vegdist
community_data <- wide_data[, c("crustacean", "echinoderms", "molluscs")]

# Remove rows where all abundance values are zero
non_zero_rows <- rowSums(community_data) > 0
community_data <- community_data[non_zero_rows, ]

# Create environmental data frame
environmental_data <- wide_data[non_zero_rows, c("TRANSECT", "YEAR", "SITE")]

@@

# Create dataframe from PERMANOVA results
permanova_df <- data.frame(
  Factor = c("SITE", "YEAR", "Residual", "Total"),
  Df = c(1, 7, 44, 52),
  SumOfSqs = c(0.5649, 5.0492, 8.7778, 14.3919),
  R2 = c(0.03925, 0.35084, 0.60991, 1.00000),
  F = c(2.8317, 3.6157, NA, NA),
  Pr_F = c(0.050, 0.001, NA, NA)  # Changed column name to Pr_F
)

@@

# Customize the flextable
flex_perm <- flex_perm %>%
  set_header_labels(
    Factor = "Factor",
    Df = "Df",
    SumOfSqs = "Sum of Squares",
    R2 = "R2",
    F = "F",
    Pr_F = "Pr(>F)"  # Changed to match the new column name
  ) %>%
  colformat_double(digits = 4) %>%
  bold(i = ~ Pr_F <= 0.05, j = "Pr_F") %>%  # Bold values if Pr_F is 0.05 or lower
  add_header_lines(values = c(
    "Permutation test for adonis under reduced model",
    "Number of permutations: 999"
  )) %>%
  autofit() %>%
  theme_box()

@@

# Perform SIMPER analysis
simper_result <- simper(community_data, group = environmental_data$SITE, permutations = 999)

@@

# Create flextable object
flex_simp <- flextable(simp)

@@

# Extract the relevant information from the SIMPER results
simper_data <- data.frame(
  Species = rownames(summary(simper_result)$`Fort Wetherill_King's Beach`),
  Contribution = summary(simper_result)$`Fort Wetherill_King's Beach`$average,
  Fort_Wetherill = summary(simper_result)$`Fort Wetherill_King's Beach`$ava,
  Kings_Beach = summary(simper_result)$`Fort Wetherill_King's Beach`$avb
)

@@

# Create the plot
simper_plot <- ggplot(simper_long, aes(x = Species, y = Abundance, fill = Site)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = sprintf("%.1f%%", Percentage[match(Species, simper_data$Species)])),
            position = position_dodge(width = 0.9),
            vjust = -0.5, 
            size = 3) +
  labs(title = "Species Contributions to Differences Between Sites",
       subtitle = "Based on SIMPER analysis",
       y = "Average Abundance",
       x = "Species") +
  scale_fill_manual(values = c("Fort_Wetherill" = "grey", "Kings_Beach" = "black")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.title = element_blank());simper_plot

# Save the plot to a file
ggsave("output/figures/simper.plot.png", plot = simper_plot, width = 15, height = 10)

@@

PERMANOVA results:
adonis2(formula = community_data ~ SITE + YEAR, data = environmental_data, permutations = 999, method = "bray")
         Df SumOfSqs      R2      F Pr(>F)    
SITE      1   0.5649 0.03925 2.8317  0.043 *  
YEAR      7   5.0492 0.35084 3.6157  0.001 ***
Residual 44   8.7778 0.60991                  
Total    52  14.3919 1.00000

SIMPER results:
ontrast: Fort Wetherill_King's Beach

            average      sd   ratio     ava     avb cumsum     p
molluscs    0.58750 0.29784 1.97250 5.57600 5.84300  0.863 0.576
echinoderms 0.07140 0.12811 0.55760 0.21400 0.34300  0.968 0.547
crustacean  0.02160 0.05018 0.43030 0.21900 0.09300  1.000 0.526
Permutation: free
Number of permutations: 999
